---
title: "hbgd"
subtitle: "healthy birth, growth & development"
author: Ryan Hafen
output:
  packagedocs:
    toc: true
navpills: |
  <li class="active"><a href='index.html'>Docs</a></li>
  <li><a href='rd.html'>Package Ref</a></li>
---

```{r global_options, include=FALSE}
# R output pre blocks are styled by default to indicate output
knitr::opts_chunk$set(comment = NA)

# shorthand for rd_link() - see ?packagedocs::rd_link for more information
rdl <- function(x) packagedocs::rd_link(deparse(substitute(x)))
```

# Intro #

The "hbgd" R package is a collection of visual and analytic methods for the study of healthy birth, growth, and development in children.

This document provides a walk through of much of the package's functionality, provided through examples and discussion.  The [package reference](rd.html) page has an exhaustive listing of all of the exported functions in the package along with detailed documentation about their parameters and usage.

## Installation ##

The hbgd package is not yet on CRAN but can be installed from a CRAN-like repository with the following:

```r
options(repos = c(tessera = "http://packages.tessera.io", getOption("repos")))
install.packages("hbgd")
```

Once the package is is installed, you simply need to load it:

```{r load, eval=TRUE, echo=TRUE}
library(hbgd)
```

# Working With Data #

The primary type of data this package is meant to deal with is longitudinal anthropometric growth measurements with associated variables having to do with health/interventions/insults, along with per-subject meta data.

## CPP sample data ##

A subset of growth data from the [collaborative perinatal project (CPP)](https://catalog.archives.gov/id/606622) is shipped with this package to serve as an example for the various methods available in the package.

The data set looks like this:

```{r cpp1, eval=TRUE, echo=TRUE, message=FALSE}
head(cpp)
```

The column names are illustrative of "standards" that the package is designed to deal with, for example `subjid` for the subject ID, `wtkg` for weight in kilograms, `htcm` for height in centimeters, etc.

While it is not required for data sets to have these variable names, it is easiest to deal with the data when it is in this format, as these names are used to get data from the WHO growth standards, associate with variable labels / descriptions, etc.

## Variable labels ##

A full list of recognized variable names and associated labels is available in a named list that comes with the package, `hbgd_labels`, also available as a data frame, `hbgd_labels_df`:

```{r hbgd_labels, eval=TRUE, echo=TRUE, message=FALSE}
DT::datatable(hbgd_labels_df, rownames = FALSE)
```

While you can use most of the functionality of this package with data in your own format, it will be most convenient to transform your variable names to match those found `hbgd_labels`.

<!-- ## Dealing with your own data ##

The functions in this package were designed to work with a specific format of longitudinal data we are working with that has been curated and been through a QC and standardization process.  By "format", we simply mean that there is a standardized form for variable names and types.
 -->

## Dealing with longitudinal data ##

Longitudinal data by definition provides repeated measurements of one or more variables over time for the same subject.  However, it often additionally provides a set of *subject-level* variables - variables that do not change over time for each subject but provide additional demographic or other information about each subject.

For example, let's look at a few rows and columns of the `cpp` data:

```{r subject_level, eval=TRUE, echo=TRUE}
cpp[1:5, c(1, 5:13)]
```

Variables like `lencm` and `bmi` are time-varying variables, but variables like `siteid` and `sex` are subject-level - they don't change within each subject.

When dealing with longitudinal data, sometimes we want to look at just subject-level variables.  In doing so, we want to transform the data so there is only one record per subject.  There is a utility function for this, `r rdl(get_subject_data())` that returns only subject-level columns of the data, with one record per subject.

```{r get_subject_data, eval=TRUE, echo=TRUE}
head(get_subject_data(cpp))
```

Likewise, if you only want to look at the time-varying variables in the data, you can use `r rdl(get_time_data())`.

```{r get_time_data, eval=TRUE, echo=TRUE}
head(get_time_data(cpp))
```

## Meta data ##

This package has a utility function, `r rdl(get_data_attributes())`, that is used behind the scenes to compute meta data about longitudinal data sets.  This function does things like infers whether a variable is subject-level or time-varying, and computes some summary statistics that are useful to have on hand like the number of subjects, etc.  This function will attach the meta data to the data set's attributes.  We mention it here because it may be useful to know that this meta data exists.

For example:

```{r attrs, eval=TRUE, echo=TRUE, message=FALSE}
cpp <- get_data_attributes(cpp)
```

This adds an attribute list called "hbgd" to the the data, which includes a summary about the data:


```{r attrs2, eval=TRUE, echo=TRUE, message=FALSE}
str(attr(cpp, "hbgd")$var_summ)
```

This tells us which variables are subject-level and time-varying variables, and which are categorical vs. numeric.  It also adds labels and computes some summary statistics.  This information is useful for some of the summary plots that we will be making.

## Dealing with length/height

Most data sets have a variable for recumbent length measurements, `lencm`, indicating the measured length of the child up to 2 years, after which a new variable for standing height, `htcm`, becomes available.  For plotting and modeling purposes, it is most often useful to deal with both measurements in the same variable.

There is a utility function, `r rdl(fix_height())`, that will take the data and return a modified data set where by default the `htcm` variable will become the combination of both `lencm` and `htcm`.  This makes the data compatible with WHO growth standard lookups since the WHO lookups in this package reference `htcm` for both length and height.

```{r fixheight, eval=TRUE, echo=TRUE, message=FALSE}
cpp <- fix_height(cpp)
```

# Growth Standards #

A major feature of this package is a collection of utility functions for conveniently converting anthropometric measurements to z-scores or centiles (and converting z-scores / centiles to measurements) for three major growth standards:

- WHO growth standard (functions prefixed with `who_`)
- Intergrowth newborn size standard (functions prefixed with `igb_`)
- Intergrowth fetal growth standard (functions prefixed with `igpre_`)

These growth standards have previously not necessarily been easy to access inside an R analysis.  In some cases R code has been provided for making conversions with the growth standard but in a form that is difficult to embed and generalize (copying and pasting code that will be frequently used is messy will almost surely lead to errors).  Some standards are provided only through published coefficients.  The goal here is to put all the standards into a single package with a unified interface, while providing additional functionality like interpolation for regions where a standard's provided tables are sparse.

The growth standard conversion methods have been painstakingly checked for accuracy through comparisons with the standards provided by the original sources.  However, we advise caution and double checking against the original sources when results will impact decisions.  Links to the original sources can be found in the sections for each standard below.

## WHO growth standard ##

The WHO growth standard provides international standard distributions for several anthropometric measurements.

The data / methodology for the WHO growth standard functions in this package come from the WHO [child growth standard](http://www.who.int/childgrowth/software/en/) and the [5-19 years growth reference](http://www.who.int/growthref/tools/en/).

WHO growth standard conversions are available for the following pairings of variables:


x_var   | y_var                                   | span of x_var
--------| ----------------------------------------| -------------------------
agedays | wtkg (weight kg)                        | ~10 years (0 - 3682 days)
agedays | htcm (height cm)                        | ~19 years (0 - 6970 days)
agedays | bmi  (BMI kg/m2)                        | ~19 years (0 - 6970 days)
agedays | hcircm (head circumference cm)          | ~ 5 years (0 - 1856 days)
agedays | muaccm (mid upper-arm circumference cm) | ~ 5 years (0 - 1856 days)
agedays | ss (subscalpular skinfold)              | ~ 5 years (0 - 1856 days)
agedays | tsftmm (triceps skinfold thickness mm)  | ~ 5 years (0 - 1856 days)
htcm    | wtkg                                    | 120 cm


Note that the WHO provides recumbent length for 0-2 years of age, after which the standing height is provided.  Here we have merged the two into a single variable, `htcm`.

Also note that in almost every case the x variable is `agedays` - the age since birth in days.  When the desired x variable is `agedays`, a simplified set of functions has been created to convert between z-scores or centiles with the following naming convention, for example, when working with `wtkg`:

- `r rdl(who_wtkg2zscore(agedays, wtkg, sex = "Female"))`: for specified `wtkg` values at ages `agedays`, get the corresponding Female z-scores
- `r rdl(who_wtkg2centile(agedays, wtkg, sex = "Female"))`: for specified `wtkg` values at ages `agedays`, get the corresponding Female centiles
- `r rdl(who_centile2wtkg(agedays, p = 50, sex = "Female"))`: for given centiles (`p=50`, or median, as default) at ages `agedays`, get the corresponding `wtkg` values
- `r rdl(who_zscore2wtkg(agedays, z = 0, sex = "Female"))`: for given z-scores (`z=0` as default) at ages `agedays`, get the corresponding `wtkg` values

If you are working with `htcm` you can swap `wtkg` for `htcm`, etc.

As z-scores are related to centiles through a simple conversion, there is a bit of redundancy in having functions for both, but both are provided for convenience.

### Examples ###

Here are some examples:

#### Get WHO median height vs. age for first-year females

Here we get the WHO median height vs. age for females weekly for the first year:

```{r who1, eval=TRUE, echo=TRUE, message=FALSE}
x <- seq(0, 365, by = 7)
med <- who_centile2htcm(x)
plot(x, med, xlab = "age in days", ylab = "median female height (cm)")
```

Here the defaults of `sex="Female"` and `p=50` were used, giving us the median height for females at the specified ages.

#### Get WHO 75th centile height vs. age for first-year males

To get the WHO 75th centile weight value for age of males at these same time points, we can do the following:

```{r who2, eval=TRUE, echo=TRUE, message=FALSE}
q75 <- who_centile2htcm(x, p = 75, sex = "Male")
```

#### Get the WHO centile for specific age and height

Suppose we want to know the WHO centile of a female child at 2 years of age (730.5 days) who is 90 cm tall:

```{r who3, eval=TRUE, echo=TRUE, message=FALSE}
who_htcm2centile(agedays = 730.5, htcm = 90)
```

This girl's height is at the 89th centile.

#### Getting values for combinations of parameters

We can send vectors of values into any of the functions and lookups in the appropriate tables will be made for each distinct case.

<!-- We can even pass in a data frame with the `data` argument, referencing the appropriate variables by name as arguments to the function. -->

For example, suppose we want to plot the WHO 99th centile of weight for age of boys and girls for the first 4 years of life:

```{r q99, eval=TRUE, echo=TRUE, message=FALSE}
dat <- data.frame(
  x   = rep(seq(0, 1461, length = 100), 2),
  sex = rep(c("Male", "Female"), each = 100))

dat$p99 <- who_centile2wtkg(dat$x, p = 99, sex = dat$sex)

xyplot(kg2lb(p99) ~ days2years(x), groups = sex, data = dat,
  ylab = "99th percentile weight (pounds) for males",
  xlab = "age (years)", auto.key = TRUE)
```

Here we are using a convenience function, `r rdl(kg2lb())`, provided with the package to plot the weights in pounds.

#### Getting z-scores for combinations of parameters

Suppose we want to compute the height and weight-for-age z-scores (HAZ and WAZ) for each subject in the cpp data:

```{r haz, eval=TRUE, echo=TRUE, message=FALSE}
haz <- who_htcm2zscore(cpp$agedays, cpp$htcm, sex = cpp$sex)
waz <- who_wtkg2zscore(cpp$agedays, cpp$wtkg, sex = cpp$sex)
```

Since `haz` and `waz` have already provided with the data, we can check that our result from the WHO standard matches that of the provided data:

```{r haz2, eval=TRUE, echo=TRUE, message=FALSE}
comp = make.groups(
  haz = data.frame(data = cpp$haz, computed = haz),
  waz = data.frame(data = cpp$waz, computed = waz)
)
xyplot(computed ~ data | which, data = comp)
```

### Generic WHO functions ###

To utility functions described so far all for measurements against age.  To deal with the case of height-for-weight WHO calculations, and also to provide a more generic interface to accessing the WHO growth standards, the following functions are also provided:

- `r rdl(who_value2zscore(x, y, x_var = "agedays", y_var = "htcm", sex = "Female"))`
- `r rdl(who_value2centile(x, y, x_var = "agedays", y_var = "htcm", sex = "Female"))`
- `r rdl(who_centile2value(x, p = 50, x_var = "agedays", y_var = "htcm", sex = "Female"))`
- `r rdl(who_zscore2value(x, z = 0, y_var = "htcm", x_var = "agedays", sex = "Female"))`

Here both the x variable and y variable can be specified.  For example, the following are equivalent:

```{r who_generic, eval=TRUE, echo=TRUE, message=FALSE}
who_htcm2centile(agedays = 730.5, htcm = 90)
who_value2centile(x = 730.5, y = 90, x_var = "agedays", y_var = "htcm")
```

But now we can do things like computing height-for-weight centiles:

```{r who_generic2, eval=TRUE, echo=TRUE, message=FALSE}
who_value2centile(x = 90, y = 12, x_var = "htcm", y_var = "wtkg")
```

## Intergrowth newborn standard ##

The [Intergrowth newborn standard](https://intergrowth21.tghn.org/articles/international-standards-newborn-weight-length-and-head-circumference-gestational-age-and-sex-newborn-cross-sectional-study-inte/) provides standards for newborn weight, length, and head circumference by gestational age and sex from gestational age of 232 days to 300 days.

Functions for this standard have a similar naming structure as we saw for the WHO standards:

- `r rdl(igb_lencm2zscore(gagebrth, lencm, sex = "Female"))`: for specified `lencm` values at gestational ages at birth `gagebrth`, get the corresponding Female z-scores
- `r rdl(igb_lencm2centile(gagebrth, lencm, sex = "Female"))`: for specified `lencm` values at gestational ages at birth `gagebrth`, get the corresponding Female centiles
- `r rdl(igb_centile2lencm(gagebrth, p = 50, sex = "Female"))`: for given centiles (`p=50`, or median, as default) at gestational ages at birth `gagebrth`, get the corresponding `lencm` values
- `r rdl(igb_zscore2lencm(gagebrth, z = 0, sex = "Female"))`: for given z-scores (`z=0` as default) at gestational ages at birth `gagebrth`, get the corresponding `lencm` values

The same functions are available for weight `wtkg` and head circumference `hcircm`.

Usage is also the same as in WHO.

#### Examples

Suppose we want to get the birth length z-scores for all subjects in our `cpp` data set:

```{r igb1, eval=TRUE, echo=TRUE}
hbaz <- igb_lencm2zscore(cpp$gagebrth, cpp$birthlen, sex = cpp$sex)
```

Also, we can make a quick and crud recreation of the boys charts seen [here](https://intergrowth21.tghn.org/site_media/media/articles/INTERGROWTH-21st_Newborn_Size_at_Birth_Chart_Boys.pdf) with this:

```{r igb2, eval=TRUE, echo=TRUE, fig.height=7}
chartdat <- do.call(rbind, lapply(c(3, 10, 50, 90, 97), function(p) {
  data.frame(p = p, gagebrth = 232:300,
  make.groups(
    wtkg = igb_centile2wtkg(232:300, p, sex = "Male"),
    lencm = igb_centile2lencm(232:300, p, sex = "Male"),
    hcircm = igb_centile2hcircm(232:300, p, sex = "Male")
  ))
}))

xyplot(data ~ gagebrth / 7 | which, groups = p, data = chartdat,
  type = c("l", "g"),
  strip = FALSE, strip.left = TRUE,
  scales = list(y = list(relation = "free")),
  layout = c(1, 3), as.table = TRUE,
  xlab = "Gestational Age at Birth (weeks)", ylab = ""
)
```

## Intergrowth prenatal standard ##

This package also has functions for dealing with the Intergrowth international standards for fetal growth based on serial ultrasound measurements.  This data covers gestational ages from 7 to 40 weeks (98 to 280 days), for the following variables:

variable   | description
-----------| --------------------------------
hccm       | head circumference (cm)
bpdcm      | biparietel diameter (cm)
ofdcm      | occipito-frontal diameter (cm)
accm       | abdominal circumference (cm)
flcm       | femur length (cm)

As with previous growth standard methods, a similar interface to this growth standard is provided with the following conventions for, for example, `hccm`:

- `r rdl(igpre_flcm2zscore(gagedays, flcm, sex = "Female"))`: for specified `flcm` values at gestational ages `gagedays`, get the corresponding Female z-scores
- `r rdl(igpre_flcm2centile(gagedays, flcm, sex = "Female"))`: for specified `flcm` values at gestational ages `gagedays`, get the corresponding Female centiles
- `r rdl(igpre_centile2flcm(gagedays, p = 50, sex = "Female"))`: for given centiles (`p=50`, or median, as default) at gestational ages `gagedays`, get the corresponding `flcm` values
- `r rdl(igpre_zscore2flcm(gagedays, z = 0, sex = "Female"))`: for given z-scores (`z=0` as default) at gestational ages `gagedays`, get the corresponding `flcm` values

Usage is similar as the other growth standards.  For example, to get the centile for child at 100 gestational days with an ultrasound head circumference measurement of 11cm:

```{r igpre1, eval=TRUE, echo=TRUE}
igpre_hccm2centile(100, 11)
```

## Plot utilities ##

This package provides a few utility functions for making it easy to add any of the growth standards as an overlay on a plot.  Methods are available for ggplot, lattice, and rbokeh.  These functions have similar parameters to the growth standard lookup functions described above.  The methods available are:

- ggplot2
  + `r rdl(geom_who())`
  + `r rdl(geom_igb())`
  + `r rdl(geom_igpre())`
- lattice
  + `r rdl(panel.who())`
  + `r rdl(panel.igb())`
  + `r rdl(panel.igpre())`
- rbokeh
  + `r rdl(ly_who())`
  + `r rdl(ly_igb())`
  + `r rdl(ly_igpre())`

Some examples are below.

#### ggplot

Here is an example of using the utility function, `r rdl(geom_who())`, with a plot of a CPP subject's height vs. age, plotting the default of WHO bands at centiles 1, 5, 25, 50, 75, 95, 99:

```{r who_gg, eval=TRUE, echo=TRUE, message=FALSE}
# make sure lencm is also in htcm before plotting
p <- ggplot(data = subset(cpp, subjid == 8), aes(x = agedays, y = htcm))
geom_who(p, x = seq(0, 2600, by = 10)) +
  geom_point()
```

#### lattice

This example of uses `r rdl(panel.who())` to superpose the WHO growth standard with a plot of a CPP subject's height vs. age, with WHO bands at z-scores -3, -2, -1, 0, 1, 2, 3:

```{r who_lattice, eval=TRUE, echo=TRUE, message=FALSE}
xyplot(wtkg ~ agedays, data = subset(cpp, subjid == 8),
  panel = function(x, y, ...) {
    panel.who(x = seq(0, 2600, by = 10),
      sex = "Male", y_var = "wtkg", p = pnorm(-3:0) * 100)
    panel.xyplot(x, y, ...)
  },
  ylim = c(0, 39),
  col = "black")
```

#### rbokeh

Here's an example plotting the same plot as the lattice plot, transforming `agedays` to age in years and taking advantage of several of rbokeh's interactive features, such as hover and zoom/pan.

```{r who_bokeh, eval=TRUE, echo=TRUE, message=FALSE}
library(rbokeh)
figure(xlab = "Age (years)", ylab = "wgkt") %>%
  ly_who(x = seq(0, 2600, by = 10), y_var = "wtkg",
    sex = "Male", x_trans = days2years) %>%
  ly_points(days2years(agedays), wtkg,
    data = subset(cpp, subjid == 8), col = "black",
    hover = c(agedays, wtkg, lencm, htcm, bmi, geniq, sysbp, diabp))
```

Look at Male birth lengths superposed on the Intergrowth birth standard:

```{r igb_bokeh, eval=TRUE, echo=TRUE, message = FALSE}
# first we need just 1 record per subject with subject-level data
cppsubj <- get_subject_data(cpp)
figure(xlab = "Gestational Age at Birth (days)", ylab = "Birth Length (cm)") %>%
  ly_igb(gagebrth = 250:310, var = "lencm", sex = "Male") %>%
  ly_points(jitter(gagebrth), birthlen, data = subset(cppsubj, sex == "Male"),
    color = "black")
```

Plot Intergrowth prenatal growth standard bands at z=1,2,3 for prenatal head circumference

```{r igpre_bokeh, eval=TRUE, echo=TRUE, message = FALSE}
figure(xlab = "Gestational Age (days)",
  ylab = "Head Circumference (cm)") %>%
    ly_igpre(gagedays = 98:280, var = "hccm", p = pnorm(-3:0) * 100)
```

In a future section we will see a utility plot function that plots a subject's fitted growth trajectory and makes use of these plot utilities.

# Simple Exploration #

Several functions are available for making simple interactive exploratory plots via the rbokeh package, and many more are planned.

Many of the summary plots have a distinction between plotting "time-varying" and "subject-level" variables.  Time-varying variables are variables that vary over time within each subject, while subject-level variables are constant within each subject.  When dealing with subject-level variables, instead of plotting summaries over the entire longitudinal data set, we instead subset the data to have one record per subject.  For example, in our CPP sample data, there are 500 subjects but nearly 2000 records.  When plotting time-varying variables, we have all 2000 records, but when plotting subject-level variables we have 500 records.

## Data distributions ##

A simple function is available to plot the univariate distributions of all variables in the data, `r rdl(plot_univar())`.  You can have it plot using subject-level or time-varying variables.  The function returns bar charts for categorical variables and histograms for quantitative variables.  The rule for deciding whether a variable is quantitative or categorical is that it is quantitative if it is numeric and has more than 10 unique values and categorical otherwise.

#### Subject-level variable summaries

```{r datadist1, eval=TRUE, echo=TRUE, message=FALSE}
plot_univar(cpp, subject = TRUE, width = 250, height = 250)
```

#### Time-varying variable summaries

```{r datadist2, eval=TRUE, echo=TRUE, message=FALSE}
plot_univar(cpp, subject = FALSE, width = 250, height = 250)
```

## Missing data ##

There are a few plot methods available for investigating missing data.  We can plot a stacked bar chart indicating NAs for each variable in a data set:

#### Subject-level variables missing values

```{r explore1, eval=TRUE, echo=TRUE, message=FALSE}
plot_missing(cpp, subject = TRUE)
```

#### Time-varying variables missing values

```{r explore1.1, eval=TRUE, echo=TRUE, message=FALSE}
plot_missing(cpp)
```

There are a lot of missing values for time-varying variables.  The only one that is complete is `wtkg`.

We can also plot a heat map of frequency of "complete" (both non-NA) pairs of variables.  The color coding ranges from light yellow for more complete cases to dark red for fewer complete cases:

#### Subject-level variables complete pairs

```{r explore2, eval=TRUE, echo=TRUE, message=FALSE}
plot_complete_pairs(cpp, subject = TRUE)
```

#### Time-varying variables complete pairs

```{r explore2.1, eval=TRUE, echo=TRUE, message=FALSE}
plot_complete_pairs(cpp)
```

## Special plots ##

There are some additional plot methods more specific to the subject matter of longitudinal data.

#### Number of visits per subject

One of these is looking at the distribution of the number of "visits" recorded for each subject.  This plots both a histogram and a quantile plot of the number of records found in the data for each subject ID.  The histogram is useful for seeing the shape of the distribution, while the quantile plot lets us see more detail.

```{r explore3, eval=TRUE, echo=TRUE, message=FALSE}
plot_visit_distn(cpp, width = 350, height = 350)
```

From this we see that a significant proportion of subjects had 4 visits, while some had as few as 1 and as many as 5.

#### Age at first visit

```{r explore4, eval=TRUE, echo=TRUE, message=FALSE}
plot_first_visit_age(cpp, width = 350, height = 350)
```

Since the observations in this data have been snapped to only a set of specific days, all subjects had their first visit on day 1.  This plot would look more interesting for other data sets where the `agedays` variable is more fluid.

#### Age frequency

Another useful plot is the frequency of `agedays` to get a feel for how many observations there are at each day of the study.

```{r explore5, eval=TRUE, echo=TRUE, message=FALSE}
agefreq <- get_agefreq(cpp)
plot_agefreq(agefreq)
```

Here we see that there are only 5 distinct values of `agedays` for which there is data recorded.

<!-- # Subject Profile Plots #

We would like to add in more plot methods that provide greater detail about a single subject in a visually effective way.  We envision a function, `subject_profile()`, which plots all variables for a given subject in a plot or provides a collection of plots that helps us understand everything about the subject.
 -->

# Modeling Methods #

An important necessity for more meaningful analyses beyond summaries or looking at the raw data is to fit models to individual subject's growth trajectories and investigate the results.

There are many reasons we may want to fit growth trajectories:

- estimate the trajectory of every subject at a specific age, when the subjects are measured at irregular time points
- study the velocity of each subject's growth
- identify data that deviates from what we perceive as regular growth (noisy around the fitted trajectory, large jumps, etc.)
- visualization

This package provides a few modeling techniques and tries to find good defaults when applied across many subjects.  The supported techniques include robust linear models via `rlm()`, robust loess smoothing via `loess()`, and splines via `gam()`, and the ability to plug in new modeling techniques is available.

## Fitting individual trajectories ##

Fitting the growth trajectory for an individual is possible using the `r rdl(fit_trajectory())` function.

To fit a trajectory, at a minimum you need to supply a data argument and the name of the y variable.

```{r fit_traj, eval=TRUE, echo=TRUE}
fit <- fit_trajectory(subset(cpp, subjid == 2), y_var = "htcm", method = "rlm")
```

Here in addition to the data and `y_var`, we also specified `method = "rlm"`.  The default trajectory fitting method is "gam", but in the case of our cpp data, the number of observations is sparse for each subject such that there are not enough points to do "gam" fitting.  You can also specify the name of the x variable if your x variable is not "agedays", which is the default.

We'll talk about some of the other parameters in a minute, but let's first look at what the output looks like.  Our `fit` object is a list with the following elements:

```{r fit_traj2, eval=TRUE, echo=TRUE}
names(fit)
```

You can investigate what each one of these looks like on your own, but here is a desription of what each of these represents:

- `xy`: a data frame with columns "x" and "y" (and optionally "z" if the variable could be converted using the WHO growth standard) of the original data being fit
- `fit`: a vector of the fitted values
- `resid`: the residuals of the fit
- `fitgrid`: a data frame with the model fits applied across an equally-spaced grid of "x" points with the following columns
  + `x`: the x-axis grid values
  + `y`: the model fits at the `x` values
  + `z`: (optional) the WHO-transformed z-scores of `y`, if y is WHO-transformable
  + `dy`: an estimate of the first-order derivative of the fit
  + `dz`: (optional) an estimate of the first-order derivative of the z-score transform of the fit, if `y` is WHO-transformable
- `checkpoint`: a data frame of "checkpoint" "x" values at which we (see the following section on trajectory checkpoints) with the following columns:
  + `x`: the age(s) at which we wish to set a checkpoint for the subject (e.g. a checkpoint at 365 and 730 days)
  + `y`: the corresponding model fit at the "x" checkpoint values
  + `z`: the z-score of the model fit at the "x" checkpoint values
  + `zcat`: a z-score categorization (e.g. <-2 or >-2) for where the subject's growth falls in the z-score scale at each checkpoint
- `pars`: a named list of fitting parameters supplied to the particular fitting method
- `data`: all columns of the data argument for this subject (preserved for later analyses)
- `sex`: the subject's sex
- `x_var`, `y_var`: the names of the variables used for "x" and "y" in the trajectory fitting

### Plotting trajectory fits

After inspecting what the output of `fit` looks like, we can get a better understanding of what's available by making some plots.  We can make simple plots by accessing the data directly:

```{r fit_plot1, eval=TRUE, echo=TRUE}
plot(fit$xy$x, fit$xy$y)
lines(fit$fitgrid$x, fit$fitgrid$y)
```

There are also several built-in plot methods.  One is simply `r rdl(plot.fittedTrajectory())`, which plots the original data points, the fitted trajectory, and the checkpoints superposed over the WHO growth standard:

```{r fit_plot2, eval=TRUE, echo=TRUE, message = FALSE}
plot(fit, who_range = c(0, 2560), hover = c("agedays", "htcm"))
```

You can specify variables to show when the raw data points are hovered.  The checkpoints are plotted as circles with x's and when hovered show the z-score category the subject falls under at that checkpoint.

You can also plot the fits on the "d-scale", which is basically taking the raw fitted curve and subtracting the WHO growth standard using `center = TRUE`:

```{r fit_plot3, eval=TRUE, echo=TRUE, message = FALSE}
plot(fit, who_range = c(0, 2560), hover = c("agedays", "htcm"), center = TRUE)
```

These are height-for-age difference (HAD) scores [see here](http://www.ifpri.org/blog/once-stunted-always-stunted-whats-catch-growth).

We may also want to plot the data on the z-score scale.  If the y-variable specified in the fitting method is z-transformable, the z-scores are automatically computed and included with the fit.  We can plot on the z-score scale with `r rdl(plot_z())`:

```{r fit_plot4, eval=TRUE, echo=TRUE, message = FALSE}
plot_z(fit, hover = c("agedays", "wtkg"))
```

Also, the fitting method automatically computes the velocity of the fit and if applicable the z-score fit.  We can plot these with:

```{r fit_plot5, eval=TRUE, echo=TRUE, message = FALSE}
plot_velocity(fit)
```

And for the z-score scale:

```{r fit_plot6, eval=TRUE, echo=TRUE, message = FALSE}
plot_zvelocity(fit)
```

There are some artificial bumps in the velocities due to jumps in the WHO standard having to do with the transition from `lencm` to `htcm` and the change in standard data at 5 years.

### Additional parameters

Additional parameters can be applied to `r rdl(fit_trajectory())` for additional control over the output, such as controlling the gridpoints for the grid fit, setting the checkpoints, setting the z-score bins for the checkpoint z-score categorizations, transformations to be applied to the x and y variables prior to fitting, and any smoothing parameters you wish to pass on to the fitting method.  See the function reference for `r rdl(fit_trajectory())` for more details.

### Fitting on the z-score scale

Often we may want to apply our trajectory fitting on the z-score scale.  We can do this by simply providing the z-score version of the variable as `y_var` (such as "waz" instead of "wtkg").  An important thing to know about this, however, is that `r rdl(fit_trajectory())` will automatically detect that this is a z-score variable and after doing the fitting, it will transform the results such that the object returned will have the same structure as those we get when we provide a raw variable for fitting.  In other words, your "y" values in the result object will be on the original scale and the "z" values in the result object will be on the z-score scale - the only difference being that the fitting was done on the z-score scale.  Here is an example:

```{r zfit, eval=TRUE, echo=TRUE}
zfit <- fit_trajectory(subset(cpp, subjid == 2), y_var = "waz", method = "rlm")
plot(zfit, hover = c("agedays", "wtkg"))
```

You can investigate the results of `zfit` and `fit` to see how they differ.

### Why use fit_trajectory

In many cases, you may want to simply run your own fitting method on the data and not use `r rdl(fit_trajectory())`.  This is completely fine.  The only things `r rdl(fit_trajectory())` offers on top of doing it yourself are the following:

- getting the fit over a grid of points for plotting
- allowing specification of growth "checkpoints" for comparisons across subjects
- dealing with fitting on the z-score scale
- providing a self-contained output with data that can be used as a foundation for other methods to build on top of (a few of which we will see later)

### Trajectory checkoints ###

The notion of computing trajectory "checkpoints" is very useful for a lot of analytical and visualization activities.

Consider the following questions:

- What percentage of children in this study were stunted at age 2?
- Is variable x different for children who were stunted (height z-score < -2) at age 1 but not stunted at age 2 vs. children who remained stunted at age 2?
- etc.

Many research questions insist that we be able to know a child's trajectory at a specific point in time.  However, most longitudinal growth data sets do not provide data at fixed time points (ages) because this is not how data is collected in real life (we cannot record every child at the exact same ages).

To estimate a child's trajectory at specific time points, we do it as part of the trajectory fitting, using the "checkpoint" argument in `r rdl(fit_trajectory())`.  We will see methods below that take the checkpoint output and help us explore differences between children that fall into different categories at different time points.

## Plugging in a new fitting method

Plugging in a new fitting method is as simple as creating a new function, `fit_method.method_name`.  For example, if I want to add a fitting method called "mymethod", I would create a function `fit_method.mymethod`.  Then it will be available to call using `fit_trajectory(..., method = "mymethod")`.  See the source code for `fit_method.loess` for an example of what inputs your method should take and what outputs it should return.

Limitations: Currently fitting trajectories is done independently for each subject.  Some methods like to see the data for all trajectories at once when fitting the data so that information can be borrowed across all the subjects.  We are working on ways to incorporate these types of fitting methods into the `r rdl(fit_trajectory())` framework.  However, this will not always be possible to accomodate, particularly in the case where we have a very large number of subjects and the only way to feasibly fit trajectories is per-subject.

## Fitting all trajectories ##

Typically in a study made up of several subjects, we want to fit growth trajectories for every subject.  A convenience function `r rdl(fit_all_trajectories())` is available to do this.

For example, to fit the weight-for-age trajectory of each subject in the `cpp` data, we can do the following:

```{r fit5, eval=TRUE, echo=TRUE}
# cppt <- fit_all_trajectories(cpp, y_var = "wtkg", method = "rlm")
```

Additional smoothing parameters can also be applied that will be passed on to the fitting method.  Here we use the "rlm" method, which probably won't do the best job fitting the trajectories but is pretty much our only choice since each subject only has 2-4 observations.

The output of `r rdl(fit_all_trajectories())` is a "distributed data object" - basically a list of trajectory fits for each subject, indexed by a key of the subject's id.  For more information about this type of data structure, see the [documentation for datadr](http://tessera.io/docs-datadr).

To inspect the output of the result of `r rdl(fit_all_trajectories())`:

```{r fitres, eval=TRUE, echo=TRUE}
# cppt
```

This tells us the result is a distributed data object with 500 subsets (we have 500 subjects).  To inspect one of the subsets:

```{r fit6, eval=TRUE, echo=TRUE}
# str(cppt[[2]], 2)
```

We see this is a list with an element "key" telling us that the data in this subset corresponds to `subjid=1`.  The "value" element should look familiar as it is simply the output of `r rdl(fit_trajectory())`.

We can plot the trajectory of this subset as we have seen before for output of `r rdl(fit_trajectory())`:

```{r fit7, eval=TRUE, echo=TRUE}
# plot(cppt[[2]]$value, hover = c("agedays", "wtkg"))
```

Note that we can also access subsets by key:

```{r fit8, eval=TRUE, echo=TRUE}
# plot(cppt[["subjid=24"]]$value, hover = c("agedays", "wtkg"))
```

We will see how to plot all of these trajectories and make them available in an interactive viewer in the upcoming [Trelliscope](#trelliscope) section.

# Division Methods #

TODO

# Trelliscope #

TODO

```{r trscope, eval=TRUE, echo=TRUE}
# trscope_trajectories(cppt)
```



<!-- Integrate model outputs from nonmem? -->


<!-- # Checkpoint Comparisons # -->


# Multi-Dataset Vis #

Everything we have seen thus far has been applied to a single data set.  This package also has some visualization methods for making visual comparisons between data sets, with plans to add more.

```{r multidat, eval=TRUE, echo=TRUE, message=FALSE}
dat_list <- list(
  cpp1 = cpp[,c(1:5, 7:9, 14:19, 23:32)],
  cpp2 = cpp[,c(1:5, 11:24)],
  cpp3 = cpp
)
```

#### Plot matrix comparing variables present in a list of studies

We can make a plot of study vs. variable color-coded by variable type with the following:

```{r multivarmat, eval=TRUE, echo=TRUE, message=FALSE}
plot_var_matrix(dat_list)
```

Variables on the x-axis are ordered from most prevalent across data sets to least, and within each group are ordered by variable type and then alphabetically.

This plot is designed to scale to a large number of data sets by growing in height for each new data set and controlling the width and making the x-axis scrollable (click and drag side to side).  Tooltips provide additional information about each study/variable combination.

#### Plot counts by age for a list of studies

```{r multicount, eval=TRUE, echo=TRUE, message=FALSE}
plot_time_count_grid(dat_list)
```

#### Compare distributions of visits per subject

```{r multivisit, eval=TRUE, echo=TRUE, message=FALSE}
plot_multi_subj_boxplot(dat_list)
```

In this case since the three data sets are the same (other than some having more columns than the other), the distributions are of course the same.

More multi-study visualization methods will be added as deemed necessary.

# Misc #

## Unit conversion ##

A few unit conversion functions are available for going between inches/centimeters, pounds/kilograms, etc.  For a list of all of these, see the help page for `r rdl(in2cm())`.

```{r whoex, eval=TRUE, echo=TRUE, message=FALSE}
who_value2zscore(1670, in2cm(44))
who_value2zscore(1670, lb2kg(48), y_var = "wtkg")
```

<!-- ## Derivative ##
`grid_deriv()`
 -->


